Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2017-05-03 manuscript.tex
latest

Here we describe how Grime's ``disturbed'', ``stressful'' and ``ideal'' environments can be captured mathematically in our scheme. To proceed, we  map these verbally-defined environments to quantitative parameter regimes in our model.

@@

Disturbed environments are characterized by short bursts of high extrinsic mortality $d$ caused by physical destruction. We assume that disturbances are equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by type $i$ each iteration. In the heavily disturbed extreme of Grime's triangle, many such bursts (which could each be spatially localized) will occur each iteration (i.e. over the time required for a propagule to reach reproductive maturity), which we approximate as almost all adults and juveniles dying every iteration $d\approx 1$. Population persistence is only possible if $(1-d)b>d$, and so $b\gg 1$ (Appendix C). We assume that fulfilling this requirement presents such a challenge that $(1-d)b$ does not greatly exceed $d$, which implies $L\ll 1$.

@@

\begin{figure}
\centering
\begin{tabular}{l*{4}{c}}
  & Ideal & Disturbance* & Stress (G) & Stress (HD) \\ \hline
  Constraints & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
  Other parameters & $b/d\gg 1$ & $(1-d)b\approx d$ & $b\approx d$ & $b/d\gg 1$ \\
  Density $L$  & $\gg 1$ & $\ll 1$ & $\ll 1$ & $\gg 1$
\end{tabular}
\caption{\label{fig:table} The mathematical realization of Grime's three environmental extremes in our model, as well as the high-density variant of the stressful environment. *Mortality affects both adults and juveniles in the disturbed environment, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$.}
\end{figure}

Given our mathematical representation of Grime's environmental extremes, summarized in Fig.~\ref{fig:table}, the expected direction of evolution is determined by three factors. First, the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}. By comparison, neutral mutations rarely fix (probability of order $1/N$). Thus $b$, $c$ and $d$ will tend to change in the direction of greatest fitness increase. Under Eq. \eqref{eq:master}, proportional changes of the form $b_j=(1+\epsilon) b_i$ or $d_j=(1+\epsilon) d_i$ have identical fitness effects

When propagule density is low $L\ll 1$, $c$ does not contribute to fitness.

Second, the fact that some mutations are more common than others can change the direction of evolution \citep{yampolsky_2001}, and in particular, very few mutations can perform difficult trait changes such as increasing $b$ under stress. These mutational constraints are different under different environments (Fig.~\ref{fig:table}).

Finally, there may be pleiotropy between $b$, $c$, and $d$, which could be due to trade-offs between them. Here we do not attempt to account for the complications posed by pleiotropy, and treat $b$, $c$ and $d$ independently.

@@

\begin{figure}
\centering
\includegraphics[scale=1]{axes.pdf}
\caption{\label{fig:axes} The realization of Grime's triangle in our model. Schematic representation of the triangular space bounded by the low/high extremes of stress/disturbance. The low-$T$ interpretation of stress is also shown. The vertices of the triangles correspond to different environmental extremes. Selection favors different traits at each vertex, leading to different trait clusters.} 
\end{figure}

>>>>

Here we describe how Grime's ``disturbed'', ``stressful'' and ``ideal'' environments can be captured mathematically in our model, and evaluate the direction of evolution for each. To proceed, we map these verbally-defined environments to quantitative parameter regimes in our model, summarized in Fig.~\ref{fig:table}.

@@

Disturbed environments are characterized by short bursts of high extrinsic mortality $d$ caused by physical destruction. We assume that disturbances are equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ (rather than $\Delta_+ n_i$) territories are secured by type $i$ each iteration. In the heavily disturbed extreme of Grime's triangle, many such bursts (which could each be spatially localized) will occur each iteration (i.e. over the time required for a propagule to reach reproductive maturity), which we approximate as a high constant mortality rate $d\approx 1$. Population persistence requires $b>d/(1-d)\gg 1$ (Appendix C), i.e. severe disturbance comes with a high fecundity/dispersal requirement. We expect that it is not possible to greatly exceed this requirement ($b\approx d/(1-d)$), which implies $L\ll 1$.

@@

\begin{figure}
\centering
\begin{tabular}{l*{4}{c}}
  & Ideal & Disturbance* & Stress (G) & Stress (HD) \\ \hline
  Constraints & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
  Other parameters & $b/d\gg 1$ & $(1-d)b\approx d$ & $b\approx d$ & $b/d\gg 1$ \\
  Propagule density $L$  & $\gg 1$ & $\ll 1$ & $\ll 1$ & $\gg 1$
\end{tabular}
\caption{\label{fig:table} The mathematical realization of Grime's three environmental extremes in our model, as well as the high-density (HD) variant of the stressful environment. *Mortality affects both adults and juveniles in the disturbed environment, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$.}
\end{figure}

The expected direction of evolution is determined by three factors. First, there may be pleiotropy between $b$, $c$, and $d$, which could be due to trade-offs between them. Here we do not attempt to account for the complications posed by pleiotropy, and treat $b$, $c$ and $d$ independently.

Second, some mutations are more common than others, which can change the direction of evolution \citep{yampolsky_2001}. In particular, very few mutations can perform difficult trait changes such as increasing resilience to intense physical disturbance. Quantifying which trait changes are more mutationally accessible than others is beyond the scope of this article, and so we will assume that $b$, $c$ and $d$ all have the same distribution of beneficial fitness effects.

Finally, the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}. Very few neutral mutations fix (probability of order $1/N$). Thus $b$, $c$ and $d$ will tend to change in the direction of greatest fitness increase.

Under Eq. \eqref{eq:master}, proportional $b$- and $d$-mutations of the form $b_j=(1+\epsilon) b_i$ and $d_j=(1-\epsilon) d_i$ have identical fitness effects (here $j$ is the mutant type; Appendix C). When propagule density is high  $L\gg 1$, the same is true for $c$-mutations of the form $c_j=(1+\epsilon) c_i$. When propagule density is low $L\ll 1$, $c$ does not contribute to fitness and $c$-mutations are neutral. Thus, in ideal and stressed environments, selection under Eq. \eqref{eq:master} does not intrinsically favor one trait improvement over another, apart from the irrelevance of $c$ at low density.

The situation in disturbed environments is more complicated. The bulk of the high mortality $d\approx 1$ is environmental and unavoidable. Thus, mutational changes in $d$ are intrinsically less effective than mutation change in $b$. However, this effect is easily swamped by the fact that disturbance mortality affects both juvenile recruitment and adult mortality.

@@

\begin{figure}
\centering
\includegraphics[scale=1]{axes.pdf}
\caption{\label{fig:axes} The realization of Grime's triangle in our model. Schematic representation of the triangular space bounded by the low/high extremes of stress/disturbance. The high-density (HD) interpretation of stress is also shown. The vertices of the triangles correspond to different environmental extremes. Selection favors different traits at each vertex, leading to different trait clusters.} 
\end{figure}
